---
title: "ACIProject"
author: "Mandy Simpson"
date: "18/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Predicting income from census data

## Introduction

The project is part of the HarvardX Profeessional Certificate in Data Science Course, PH:125.9x - Capstone.

For the self-directoed project I chose to explore the Adult Census Income dataset ("ACI") from University of California at Irvine, which can be found at <https://www.kaggle.com/uciml/adult-census-income>.  This dataset was originally derived from the US Census Bureau 1994 database.  It includes an indicator column showing whether the individual described by that entry earned under or over $50k. It is this indicator that I am seeking to predict using the other information in the dataset. 

I started the project by exploring the full dataset to ensure I understood the information contained. While the data was already broadly clean, and in tidy format, I removed some rows with missing data, and stratified the data in areas such as age, and education level. 

Once this was complete, I split the dataset into training and validation sets (90%/10%), and then split the training set further into training and test sets (80%/20%), to allow a number of algorithms to be trialled on this data before selecting one for use on the validation set. 

#Downloading and exploring the Adult Census Income dataset

Code for downloading!!
```{r download, message=FALSE, warning=FALSE}
adult_census_income <- read.csv("adult.csv")
```

The ACI dataset as downloaded has `r nrow(adult_census_income)` observations each representing an individual person in the census. 

There are `r ncol(adult_census_income) attributes as follows:

Name               | Type                | Description                               |
-------------------|---------------------|-------------------------------------------|
age                | Integer             | Age of respondent                         |
workclass          | Factor (9 levels)   | Employer type                             |
fnlwgt             | Integer             | Population weighting factor               |
education          | Factor (16 levels)  | Highest education level achieved          |
education.num      | Integer             | Numerical representation of education     |
marital.status     | Factor (7 levels)   | Marital Status of respondent              |
occupation         | Factor (15 levels)  | Type of employment                        |
relationship       | Factor (6 levels)   | Position in household                     |
race               | Factor (5 levels)   | Race of respondent                        |
sex                | Factor (2 levels)   | Sex of respondent                         |
capital.gain       | Integer             | Investment gains                          |
capital.loss       | Integer             | Investment losses                         |
hours.per.week     | Integer             | Working hours                             |
native.country     | Factor (42 levels)  | Country of birth                          |
income             | Factor (2 levels)   | Income under or over $50k                 |



